---
title: "Clinical Data Exploration"
author: "Carlos Ruiz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

In this document, I will explore the clinical variables present in the IPPSmol dataset.

# Data loading

```{r}
library(tidyverse)
library(GGally)
library(survival)
library(survminer)
library(cowplot)
```

```{r}
load("../results/preprocess/clinical_preproc.Rdata")
```

```{r}
colnames(clinical_all)
```

The dataset contains `r nrow(clinical_all)` individuals and `r ncol(clinical_all)` variables.

# Data exploration

Define functions

```{r}
plot_continuous <- function(var, binwidth){
  ggplot(clinical_all, aes(x = .data[[var]])) +
    geom_histogram(binwidth = binwidth) +
    theme_bw()
}

descrip_continuous <- function(var, thres){
  clinical_all %>%
    mutate(intervals = cut_interval(.data[[var]], thres)) %>%
    count(intervals) %>%
    mutate(perc = round(n/sum(n)*100, 1))
}


plot_categorical <- function(var){
  ggplot(clinical_all, aes(x = .data[[var]])) +
    geom_bar() +
    theme_bw()
}

descrip_categorical <- function(var){
  clinical_all %>%
    count(.data[[var]]) %>%
    mutate(perc = round(n/sum(n)*100, 1))
}
```


## Demographic variables

The cohort contains two demographic variables: age and sex.

```{r}
demo_vars <- c("AGE", "SEX")
```


```{r}
plot_continuous("AGE", 1)
descrip_continuous("AGE", 10)
```

Most individuals are between 56 and 90 years old. Nonetheless, `r round(mean(clinical_all$AGE < 56, na.rm = TRUE)*100, 2)`% of the individuals are younger than 56, being `r round(mean(clinical_all$AGE < 30, na.rm = TRUE)*100, 2)`% younger than 30. In addition, `r round(mean(is.na(clinical_all$AGE))*100, 2)`% have missing age.

```{r}
plot_categorical("SEX")
descrip_categorical("SEX")
```

There are more men than woman. 

```{r}
ggplot(clinical_all, aes(x = SEX, y = AGE)) +
  geom_boxplot() +
  theme_bw()

ggplot(clinical_all, aes(x = AGE, color = SEX)) +
 geom_freqpoly( binwidth = 1) + theme_bw()
```

The age distribution seems very similar in men than women

## Clinical variables

```{r}
clin_vars <- c("BM_BLAST", "PB_BLAST", "WBC", "ANC", "MONOCYTES", "HB", "PLT")
```

### Bone marrow blasts (%)

```{r}
plot_continuous("BM_BLAST", 1)

descrip_continuous("BM_BLAST", 10)
```

Bone marrow blasts show a tail distribution, with most individuals between 0 and 5% but a tail up to 20%.

### Peripheral blood blasts (%)

```{r}
plot_continuous("PB_BLAST", 1)

descrip_continuous("PB_BLAST", 10)
```

Most individuals did not have blasts in peripheral blood. A high proportion of samples (`r round(mean(is.na(clinical_all$PB_BLAST))*100, 2)`%) had missing values.


### Whole blood cell counts (Giga/L)

```{r}
plot_continuous("WBC", 1)

descrip_continuous("WBC", 10)
```

Most samples had white blood cell counts between 1.6 and 6.6 Giga/L. A relevant proportion of samples (`r round(mean(is.na(clinical_all$WBC))*100, 2)`%) had missing values for white blood cell counts.

### Absolute Neutrophil Count (Giga/L)

```{r}
plot_continuous("ANC", 1)

descrip_continuous("ANC", 10)
```

Most samples had neutrophil counts between 0 and 3 (Giga/L). Some samples (`r round(mean(is.na(clinical_all$WBC))*100, 2)`%) had missing values for neutrophil counts.

### Monocyte Count (Giga/L)

```{r}
plot_continuous("MONOCYTES", 1)

descrip_continuous("MONOCYTES", 10)
```

Most samples had monocyte counts below 1 Giga/L. A high proportion of sample (`r round(mean(is.na(clinical_all$WBC))*100, 2)`%) had missing values for monocyte counts.


### Hemoglobin (g/dL)

```{r}
plot_continuous("HB", 1)

descrip_continuous("HB", 10)
```

Most samples had hemoglobin proportions between 8 and 12 g/dL. 

### Platelets (Giga/L)

```{r}
plot_continuous("PLT", 10)
plot_continuous("PLT", 1) +
  xlim(c(0, 250))

descrip_continuous("PLT", 10)
```

Most samples had platelet count values below 200 Giga/L. Original studied capped the platelet values to 250. 

### Ringed Sideroblasts (Giga/L)

```{r}
plot_continuous("RINGED_SIDEROBLASTS", 1)

descrip_continuous("RINGED_SIDEROBLASTS", 10)
```

We assumed that individuals without ringed sideroblasts data did not have this event.

## Cytogenetics

```{r}
kar_events <- c("monoy", "del11q", "del5q", "del12p",
                "del20q", "del7q", "triso8", "triso19", "mono7", 
                "i17q", "chr3ab", "any_aberration")

cyto <- c("CYTO_IPSSR", "complex", kar_events)
```

### Cytogenetics IPSSR

```{r}
plot_categorical("CYTO_IPSSR")
descrip_categorical("CYTO_IPSSR") 
```

Most of the individuals presented good or very good cytogenetic results.


### Complex

```{r}
plot_categorical("complex")
descrip_categorical("complex") 
```

Some individuals present complex karyotypes.


```{r}
ggplot(clinical_all, aes(x = complex, y = N_aberrations)) +
  geom_boxplot() + 
  theme_bw()
```
Complex karyotypes had a much higher number of aberrations. 


### Karyotype events


```{r}
clin_cyto <- clinical_all %>%
  select(kar_events) %>%
  gather(Event, Value) %>%
  filter(Value != 0)

ggplot(clin_cyto, aes(x = Event)) +
    geom_bar() +
  theme_bw()
clin_cyto %>%
  count(Event) %>%
    mutate(perc = round(n/nrow(clinical_all)*100, 1)) %>%
  arrange(desc(n))
```

Almost half of the patients had at least a karyotipic aberration. The most common  event is del5q. 

```{r}
plot_continuous("N_aberrations", 1)

descrip_continuous("N_aberrations", 10)
```

Most individuals did not have any karyotipic event. Patients with events had 1 or 2. 


## MDS subtypes

```{r}
mds_types <- c("MDS_TYPE", "WHO_2016")
```


### MDS type

```{r}
plot_categorical("MDS_TYPE")
descrip_categorical("MDS_TYPE") 
```

Most patients are primary MDS. However, a subset where secondary or therapy related MDS.


### WHO subtypes

```{r}
plot_categorical("WHO_2016") 
descrip_categorical("WHO_2016")  %>%
  arrange(desc(n))
```

Most frequent categories are MDS-MLD or MDS with excess of blasts (EB1 and/or EB2)

## IPSS scores

```{r}
scores <- c("IPSSR", "IPSSR_SCORE", "IPSSRA", "IPSSRA_SCORE", "IPSSM", "IPSSM_SCORE")
```


### IPSSR

```{r}
plot_continuous("IPSSR_SCORE", 0.5)
plot_categorical("IPSSR") 
descrip_categorical("IPSSR") 
```

Individuals are more likely to have low IPSSR scores. For low scores, there is a tendency to overestimate entire scores (i.e. 1, 2,...)


### IPSSRA

```{r}
plot_continuous("IPSSRA_SCORE", 0.5)
plot_categorical("IPSSRA") 
descrip_categorical("IPSSRA") 
```

The IPSSRA scores show a more normal distribution. Nonetheless, it still predominates the individuals with low scores.


### IPSSM

```{r}
plot_continuous("IPSSM_SCORE", 0.5)
plot_categorical("IPSSM") 
descrip_categorical("IPSSM") 
```

The IPSSM scores show also negative values, reflecting individuals with lower risk than expected. Nonetheless, it still biased to low risk individuals.

## Outcomes

```{r}
outcomes <- c("OS_YEARS", "OS_STATUS", "AMLt_YEARS", "AMLt_STATUS", "LFS_YEARS", "LFS_STATUS" )
```

### Overall Survival

```{r}
os_surv <- survfit(Surv(OS_YEARS,OS_STATUS) ~ 1, clinical_all)
print(os_surv)
ggsurvplot(os_surv, data = clinical_all) 
```

Almost half of the individuals died during the follow-up. Survival probability decreased sharply and stabilizes around 5 years.

### Acute Myeloid Leukemia transformation; 

```{r}
aml_surv <- survfit(Surv(AMLt_YEARS,AMLt_STATUS) ~ 1, clinical_all)
print(aml_surv)
ggsurvplot(aml_surv, data = clinical_all) 
```

A small proportion of patients developed Acute Myeloid Leukemia transformation. Transformation happened relative early. 


### Leukemia-Free Survival

```{r}
lfs_surv <- survfit(Surv(LFS_YEARS,LFS_STATUS) ~ 1, clinical_all)
print(lfs_surv)
ggsurvplot(lfs_surv, data = clinical_all) 
```

A similar proportion of patients that died developed leukemia. The decrease is sharper and peaks around 3 years after starting the follow-up.

## Treatments

```{r}
treatments <- c("hma", "lenalidomid", "chemotherapy", "transplant")
```


```{r}
clinical_all <- clinical_all %>%
  mutate(N_treatments = hma + lenalidomid + chemotherapy + transplant) 


plot_categorical("N_treatments") 
descrip_categorical("N_treatments") 
```

Most patients did not have any treatment. Of the patients treated, most have just one treatment.

```{r}
clin_sum <- clinical_all %>%
  select(hma, lenalidomid, chemotherapy, transplant) %>%
  gather(Treatment, Value) %>%
  filter(Value != 0)

ggplot(clin_sum, aes(x = Treatment)) +
    geom_bar() +
  theme_bw()
clin_sum %>%
  count(Treatment)
```

The most frequent treatment was hypomethylation agents (hma).

# Variables covariation

## Demographic vs clinical

```{r}
ggpairs(clinical_all[, c("AGE", clin_vars)])
plot_grid(plotlist = 
            lapply(clin_vars, function(var){
   ggplot(clinical_all, aes(x = SEX, y = .data[[var]])) +
    geom_boxplot() +
    theme_bw()
}), nrow = 2)

summary(lm(PLT ~ SEX, clinical_all))
```

Among the clinical variables, we observed correlation between the blood cell counts (WBC, ANC and MONOCYTES). Age is not correlated with any clinical variable. Female had slightly higher platelet levels.

## Demographic vs cytogenetics

```{r}
ggpairs(clinical_all[, c(demo_vars, "CYTO_IPSSR", "complex")])
```

There is no correlation between age and sex and cytogenetics. 


```{r}
plot_grid(plotlist = 
            lapply(kar_events[1:6], function(var){
   ggplot(clinical_all, aes(color = factor(.data[[var]]), x = AGE)) +
    geom_density() +
    theme_bw()
}), nrow = 3)

plot_grid(plotlist = 
            lapply(kar_events[7:12], function(var){
   ggplot(clinical_all, aes(color = factor(.data[[var]]), x = AGE)) +
    geom_density() +
    theme_bw()
}), nrow = 3)

summary(glm(del7q ~ AGE, clinical_all, family = "binomial" ))
summary(glm(monoy ~ AGE, clinical_all, family = "binomial" ))
summary(glm(del12p ~ AGE, clinical_all, family = "binomial" ))

summary(glm(mono7 ~ AGE, clinical_all, family = "binomial" ))
summary(glm(chr3ab ~ AGE, clinical_all, family = "binomial" ))
summary(glm(triso19 ~ AGE, clinical_all, family = "binomial" ))
summary(glm(any_aberration ~ AGE, clinical_all, family = "binomial" ))


```

Karyotipic aberrations are not related with age. LOY is increased in older patients while monosomies 7 is associated with younger patients.



```{r}
plot_grid(plotlist = 
            lapply(kar_events[1:6], function(var){
   ggplot(clinical_all, aes(x = factor(.data[[var]]), fill = SEX)) +
    geom_bar(position = "dodge") +
    theme_bw()
}), nrow = 3)

plot_grid(plotlist = 
            lapply(kar_events[7:12], function(var){
   ggplot(clinical_all, aes(x = factor(.data[[var]]), fill = SEX)) +
    geom_bar(position = "dodge") +
    theme_bw()
}), nrow = 3)


fisher.test(table(clinical_all$del5q, clinical_all$SEX))
fisher.test(table(clinical_all$triso8, clinical_all$SEX))
fisher.test(table(clinical_all$any_aberration, clinical_all$SEX))
```

del5q is more frequent in women than in men, while trisomy 8 is more likely in men. Having any type of aberration is not associated with sex.  

```{r}
ggpairs(clinical_all[, c(demo_vars, "N_aberrations")])
```
There is no correlation between age or sex and the number of karyotipic aberrations. 

## Demographic vs subtypes

```{r}
ggpairs(clinical_all[, c(demo_vars, mds_types)])
```

There might be an association between age and WHO subtypes. 

## Demographic vs risks

```{r}
ggpairs(clinical_all[, c(demo_vars, "IPSSR", "IPSSRA", "IPSSM")])
```

There is no association between IPSS categories and sex or age. 

## Demographic vs survival

### Overall Survival

```{r}
clin_age <- mutate(clinical_all, age_int = cut_interval(AGE, 5))
os_age <- survfit(Surv(OS_YEARS,OS_STATUS) ~ age_int, clin_age)
ggsurvplot(os_age, data = clin_age) 
coxph(Surv(OS_YEARS,OS_STATUS) ~ AGE, clinical_all)

```
Older patients had worse survival.

```{r}
os_sex <- survfit(Surv(OS_YEARS,OS_STATUS) ~ SEX, clinical_all)
ggsurvplot(os_sex, data = clinical_all) 
coxph(Surv(OS_YEARS,OS_STATUS) ~ SEX, clinical_all)
```

Men had worse survival than women.

```{r}
os_demo <- survfit(Surv(OS_YEARS,OS_STATUS) ~ SEX + age_int, clin_age)
ggsurvplot(os_demo, data = clin_age) 
coxph(Surv(OS_YEARS,OS_STATUS) ~ SEX + AGE, clinical_all)

```

Women had higher survival at all ages.

### Acute Myeloid Leukemia transformation

```{r}
aml_age <- survfit(Surv(AMLt_YEARS,AMLt_STATUS) ~ age_int, clin_age)
ggsurvplot(aml_age, data = clin_age) 
coxph(Surv(AMLt_YEARS,AMLt_STATUS) ~ AGE, clinical_all)
```

Middle-age patients (47-80 yo) have the highest risk for Acute Myeloid Leukemia transformation.

```{r}
aml_sex <- survfit(Surv(AMLt_YEARS,AMLt_STATUS) ~ SEX, clinical_all)
ggsurvplot(aml_sex, data = clinical_all) 
coxph(Surv(AMLt_YEARS,AMLt_STATUS) ~ SEX, clinical_all)

```

There were no differences between men and women for AML transformation.

### Leukemia-Free Survival

```{r}
lfs_age <- survfit(Surv(LFS_YEARS,LFS_STATUS) ~ age_int, clin_age)
ggsurvplot(lfs_age, data = clin_age) 
coxph(Surv(LFS_YEARS,LFS_STATUS) ~ AGE, clinical_all)
```

Older patients have higher risk for developing leukemia.

```{r}
lfs_sex <- survfit(Surv(LFS_YEARS,LFS_STATUS) ~ SEX, clinical_all)
ggsurvplot(lfs_sex, data = clinical_all) 
coxph(Surv(LFS_YEARS,LFS_STATUS) ~ SEX, clinical_all)

```

Men also had higher risk for suffering leukemia.


```{r}
lfs_demo <- survfit(Surv(LFS_YEARS,LFS_STATUS) ~ SEX + age_int, clin_age)
ggsurvplot(lfs_demo, data = clin_age) 
coxph(Surv(LFS_YEARS,LFS_STATUS) ~ SEX + AGE, clinical_all)

```

The differences between men and women are consistent for all the ages.


## Clinical vs cytogenetics

```{r}
plot_grid(plotlist = 
            lapply(clin_vars, function(var){
   ggplot(clinical_all, aes(x = CYTO_IPSSR, y = .data[[var]])) +
    geom_boxplot() +
    theme_bw()
}), nrow = 2)

plot_grid(plotlist = 
            lapply(clin_vars, function(var){
   ggplot(clinical_all, aes(x = complex, y = .data[[var]])) +
    geom_boxplot() +
    theme_bw()
}), nrow = 2)
```

Cytogenetic measurements were reflected in cell types proportions.


```{r}
lapply(kar_events, function(kar_var){
  plot_grid(plotlist = 
            lapply(clin_vars, function(var){
   ggplot(clinical_all, aes(x = factor(.data[[kar_var]]), y = .data[[var]])) +
    geom_boxplot() +
    theme_bw()
  }), nrow = 2)
})
```
Cytogenetic events were reflected in cell types proportions.

```{r}
ggpairs(clinical_all[, c(clin_vars, "N_aberrations")])
```

The number of aberrations was slighlty correlated with clinical measurements.

## Clinical vs subtypes

```{r}
plot_grid(plotlist = 
            lapply(clin_vars, function(var){
   ggplot(clinical_all, aes(x = MDS_TYPE, y = .data[[var]])) +
    geom_boxplot() +
    theme_bw()
}), nrow = 2)

plot_grid(plotlist = 
            lapply(clin_vars, function(var){
   ggplot(clinical_all, aes(x = WHO_2016, y = .data[[var]])) +
    geom_boxplot() +
    theme_bw()
}), nrow = 2)

```

There is high correlation between WHO subtypes and clinical variables. Nonetheless, WHO subtypes are defined based on clinical variables.


## Clinical vs risks

```{r}
ggpairs(clinical_all[, c(clin_vars, "IPSSR_SCORE", "IPSSRA_SCORE", "IPSSM_SCORE")])
```

Clinical variables are highly correlated with scores.

## Clinical vs Survival

```{r}
coxph(formula(paste("Surv(OS_YEARS,OS_STATUS) ~",  paste(clin_vars, collapse = "+"))),
              clinical_all)
```

Blast proportions, hemoglobin and platelets are associated with overall survival.


```{r}
coxph(formula(paste("Surv(AMLt_YEARS,AMLt_STATUS) ~",  paste(clin_vars, collapse = "+"))),
              clinical_all)
```

Bone marrow blasts, whole blood counts and neutrophil counts are associated with AML transformation.

```{r}
coxph(formula(paste("Surv(LFS_YEARS,LFS_STATUS) ~",  paste(clin_vars, collapse = "+"))),
              clinical_all)
```

Bone marrow proportions, hemoglobin and platelets are associated with leukemia free survival.


## Karyotype vs Survival

### Overall survival

```{r}
coxph(Surv(OS_YEARS,OS_STATUS) ~ any_aberration, clinical_all)
```

Having any karyotype aberration is associated with survival.

```{r}
coxph(Surv(OS_YEARS,OS_STATUS) ~ N_aberrations, clinical_all)
```

The number of aberrations is associated with worse survival.

```{r}
single_kar <- kar_events[kar_events != "any_aberration"]
coxph(formula(paste("Surv(OS_YEARS,OS_STATUS) ~",  paste(single_kar, collapse = "+"))),
              clinical_all)
```

Del5q, del12p, del7q, trisomy8, monosomy7, i17q and chr3 aberrations are associated with worse overall survival.


```{r}
coxph(Surv(OS_YEARS,OS_STATUS) ~ monoy, clinical_all, subset = SEX == "M")
```
LOY is associated with better survival.

### Acute Myeloid Leukemia transformation


```{r}
coxph(Surv(AMLt_YEARS,AMLt_STATUS) ~ any_aberration, clinical_all)
```

Having any karyotype aberration is associated with AML transformation.

```{r}
coxph(Surv(AMLt_YEARS,AMLt_STATUS) ~ N_aberrations, clinical_all)
```

The number of aberrations is associated with AML transformation.

```{r}
coxph(formula(paste("Surv(AMLt_YEARS,AMLt_STATUS) ~",  paste(single_kar, collapse = "+"))),
              clinical_all)
```

Del5q, del12p, del20q, del7q, trisomy8, monosomy7 and chr3 aberrations are associated with AML transformation.


```{r}
coxph(Surv(AMLt_YEARS,AMLt_STATUS) ~ monoy, clinical_all, subset = SEX == "M")
```

LOY is protective of AMLt.

### Leukemia-Free Survival

```{r}
coxph(Surv(LFS_YEARS,LFS_STATUS) ~ any_aberration, clinical_all)
```

Having any karyotype aberration is associated with Leukemia Free Survival (LFS).

```{r}
coxph(Surv(LFS_YEARS,LFS_STATUS) ~ N_aberrations, clinical_all)
```

The number of aberrations is associated with having Leukemia.

```{r}
coxph(formula(paste("Surv(LFS_YEARS,LFS_STATUS) ~",  paste(single_kar, collapse = "+"))),
              clinical_all)
```

Del5q, del12p, del7q, trisomy8, monosomy7, i17q and chr3 aberrations are associated with LFS.


```{r}
coxph(Surv(LFS_YEARS,LFS_STATUS) ~ monoy, clinical_all, subset = SEX == "M")
```