---
title: "Main results of low-blasts clustering"
author: "Carlos Ruiz"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    toc_depth: 3
    downcute_theme: "default"
    lightbox: true
    thumbnails: true
    gallery: true
    use_bookdown: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Preparación

```{r}
library(MASS)
library(cluster)
library(rpart)
library(ca)
library(cowplot)
library(pheatmap)
library(survminer)
library(survival)
library(tidyverse)
library(DT)
library(rmdformats)
library(ggpubr)
```

```{r}
load("../results/clustering/lowblast_filter_pc_clustering.Rdata")
load("../results/clustering/lowblast_filter_pc_clustering_raw.Rdata")
load("../results/preprocess/clinical_preproc.Rdata")

```

```{r}
kar_events <- c("delY", "del11q", "del20q", "del7q",  "plus8",  "del7")
sel_muts <- c("TET2", "ASXL1", "SRSF2", "DNMT3A", "RUNX1", 
              "STAG2", "U2AF1", "EZH2", "ZRSR2", "TET2bi", "TET2other")
#clinical_low$clust <- clinical_low$clust_collapsed
clinical_low$logPLT <- log(clinical_low$PLT)
```


## Project justification

Correlation between consensus classification and IPSSM

```{r}
cons_ipss_tab <- table(clinical$consensus, clinical$IPSSM)
pilltabs(cons_ipss_tab)
```

```{r}
consensus_ipssm_prop <- prop.table(table(clinical$consensus, clinical$IPSSM), margin = 1) %>%
    as.matrix()*100 
pheatmap(consensus_ipssm_prop, display_numbers = TRUE, cluster_cols = FALSE)
```

## Data descriptives

```{r}
getIQR <- function(vec){
   sprintf("%.1f (%.1f-%.1f)", 
           median(vec, na.rm = TRUE), 
           quantile(vec, probs = 0.25, na.rm = TRUE),
           quantile(vec, probs = 0.75, na.rm = TRUE))
}
```


```{r}
descriptives <- clinical_low %>%
  group_by(clust) %>%
  summarize(N = sprintf("%i (%.1f%%)", n(), n()/nrow(clinical_low)*100),
            Females = sprintf("%i (%.1f%%)", sum(SEX == "F"), mean(SEX == "F")*100),
            Age = getIQR(AGE),
            `Bone Marrow blasts` = getIQR(BM_BLAST),
            `Whole Blood Cell count` = getIQR(WBC),
            `Neutrophil Count` = getIQR(ANC),
            HB = getIQR(HB),
            logPLT = getIQR(logPLT),
            `Y-` =  sprintf("%i (%.1f%%)", sum(delY == 1), mean(delY == 1)*100),
            `7-` =  sprintf("%i (%.1f%%)", sum(del7 == 1), mean(del7 == 1)*100),
            `8+` =  sprintf("%i (%.1f%%)", sum(plus8 == 1), mean(plus8 == 1)*100),
            del20q =  sprintf("%i (%.1f%%)", sum(del20q == 1), mean(del20q == 1)*100),
            ASXL1 =  sprintf("%i (%.1f%%)", sum(ASXL1 == 1), mean(ASXL1 == 1)*100),
            DNMT3A =  sprintf("%i (%.1f%%)", sum(DNMT3A == 1), mean(DNMT3A == 1)*100),
            RUNX1 =  sprintf("%i (%.1f%%)", sum(RUNX1 == 1), mean(RUNX1 == 1)*100),
            `IPSSM Very-Low` = sprintf("%i (%.1f%%)", sum(IPSSM == "Very-Low", na.rm = TRUE), mean(IPSSM == "Very-Low", na.rm = TRUE)*100),
            `IPSSM Low` = sprintf("%i (%.1f%%)", sum(IPSSM == "Low", na.rm = TRUE), mean(IPSSM == "Low", na.rm = TRUE)*100),
            `IPSSM Moderate-Low` = sprintf("%i (%.1f%%)", sum(IPSSM == "Moderate-Low", na.rm = TRUE), mean(IPSSM == "Moderate-Low", na.rm = TRUE)*100),
            `IPSSM Moderate-High` = sprintf("%i (%.1f%%)", sum(IPSSM == "Moderate-High", na.rm = TRUE), mean(IPSSM == "Moderate-High", na.rm = TRUE)*100),
            `IPSSM High` = sprintf("%i (%.1f%%)", sum(IPSSM == "High", na.rm = TRUE), mean(IPSSM == "High", na.rm = TRUE)*100),
            `IPSSM Very-High` = sprintf("%i (%.1f%%)", sum(IPSSM == "Very-High", na.rm = TRUE), mean(IPSSM == "Very-High", na.rm = TRUE)*100)) %>%
  rbind(tibble(clust = "Total", 
      
clinical_low %>%
  summarize(N = sprintf("%i (%.1f%%)", n(), n()/nrow(clinical_low)*100),
            Females = sprintf("%i (%.1f%%)", sum(SEX == "F"), mean(SEX == "F")*100),
            Age = getIQR(AGE),
            `Bone Marrow blasts` = getIQR(BM_BLAST),
            `Whole Blood Cell count` = getIQR(WBC),
            `Neutrophil Count` = getIQR(ANC),
            HB = getIQR(HB),
            logPLT = getIQR(logPLT),
            `Y-` =  sprintf("%i (%.1f%%)", sum(delY == 1), mean(delY == 1)*100),
            `7-` =  sprintf("%i (%.1f%%)", sum(del7 == 1), mean(del7 == 1)*100),
            `8+` =  sprintf("%i (%.1f%%)", sum(plus8 == 1), mean(plus8 == 1)*100),
            del20q =  sprintf("%i (%.1f%%)", sum(del20q == 1), mean(del20q == 1)*100),
            ASXL1 =  sprintf("%i (%.1f%%)", sum(ASXL1 == 1), mean(ASXL1 == 1)*100),
            DNMT3A =  sprintf("%i (%.1f%%)", sum(DNMT3A == 1), mean(DNMT3A == 1)*100),
            RUNX1 =  sprintf("%i (%.1f%%)", sum(RUNX1 == 1), mean(RUNX1 == 1)*100),
            `IPSSM Very-Low` = sprintf("%i (%.1f%%)", sum(IPSSM == "Very-Low", na.rm = TRUE), mean(IPSSM == "Very-Low", na.rm = TRUE)*100),
            `IPSSM Low` = sprintf("%i (%.1f%%)", sum(IPSSM == "Low", na.rm = TRUE), mean(IPSSM == "Low", na.rm = TRUE)*100),
            `IPSSM Moderate-Low` = sprintf("%i (%.1f%%)", sum(IPSSM == "Moderate-Low", na.rm = TRUE), mean(IPSSM == "Moderate-Low", na.rm = TRUE)*100),
            `IPSSM Moderate-High` = sprintf("%i (%.1f%%)", sum(IPSSM == "Moderate-High", na.rm = TRUE), mean(IPSSM == "Moderate-High", na.rm = TRUE)*100),
            `IPSSM High` = sprintf("%i (%.1f%%)", sum(IPSSM == "High", na.rm = TRUE), mean(IPSSM == "High", na.rm = TRUE)*100),
            `IPSSM Very-High` = sprintf("%i (%.1f%%)", sum(IPSSM == "Very-High", na.rm = TRUE), mean(IPSSM == "Very-High", na.rm = TRUE)*100))
         ))


descriptives_tab <- t(descriptives[, -1])
colnames(descriptives_tab) <- as.character(descriptives$clust)
datatable(descriptives_tab)
```

## Clinical variables

### Bone Marrow Blasts

```{r}
clinical_low %>%
    mutate(category = cut(BM_BLAST, breaks = c(-0.1, 1, 2, 3, 4, 5, 10, 20), 
                            labels = c("0", "1", "2", "3", "4", "5-10", "10+"))) %>%
    group_by(category, clust) %>%
    summarize(n_cat = n()) %>%
    group_by(clust) %>%
    mutate(prop = n_cat/sum(n_cat),
           category = factor(category, levels = as.character(4:0))) %>%
  ggplot(aes(x = clust, y = prop, fill = category)) +
    geom_bar(stat = "identity") +
    theme_bw() +
    labs(
    title = "Bone Marrow Blasts",
    y = "Proportion of individuals",
    x = "Cluster",
    fill = "BM (%)") +
    scale_fill_manual(values = c("#7a0177", "#c51b8a", "#f768a1", "#fa9fb5", "#fde0dd")) +
  theme(plot.title = element_text(hjust = 0.5))

```


```{r}
bm_df <-  clinical_low %>%
  mutate(cl = ifelse(clust %in% c("TET2 bi", "del7"), "High", "Other"))
bm_high <- summary(glm(BM_BLAST ~ cl, bm_df, family = "poisson"))
```

TET2bi and del7 have a higher proportion of bone marrow blasts (P = `r sprintf("%.1e", bm_high$coefficients[2, 4], scientific = TRUE)`) than the other clusters.

### Whole Blood Cell count (Giga/L)

```{r}
ggplot(clinical_low, aes(x = clust, y = WBC)) +
  geom_boxplot() +
  theme_bw() +
    labs(
    title = "Whole Blood Cells ",
    y = "WBC (Giga/L)",
    x = "Cluster") +
  theme(plot.title = element_text(hjust = 0.5))

```

```{r}
wbc_Aberr_TET2High_df <-  clinical_low %>% 
  subset(clust %in% c("8+", "del7", "del20q", "High Leukopenic",  "TET2 mono", "TET2 bi")) %>%
  mutate(cl = ifelse(clust %in% c("8+", "del7"), "aber", "Other"))
wbc_Aberr_TET2High <- summary(glm(WBC ~ cl, wbc_Aberr_TET2High_df, family = "poisson"))

wbc_TET2High_Y_df <-  clinical_low %>% 
  subset(clust %in% c("del20q", "High Leukopenic", "TET2 bi", "TET2 mono", "Y-")) %>%
  mutate(cl = ifelse(clust %in% c("Y-"), "Y-", "Other"))
wbc_TET2High_Y <- summary(glm(WBC ~ cl, wbc_TET2High_Y_df, family = "poisson"))

wbc_Y_Mild_df <-  clinical_low %>% 
  subset(clust %in% c("Mild Leukopenic", "Y-")) %>%
  mutate(cl = ifelse(clust == "Y-", "Y-", "Other"))
wbc_Y_Mild <- summary(glm(WBC ~ cl, wbc_Y_Mild_df, family = "poisson"))
```

Cluster are ordered as follows:

8+, del7 < High Leukopenic, del20q, TET2 bi, TET2 mono (P = `r sprintf("%.1e", wbc_Aberr_TET2High$coefficients[2, 4], scientific = TRUE)`)

High Leukopenic, del20q, TET2 bi, TET2 mono < Y- (P = `r sprintf("%.1e", wbc_TET2High_Y$coefficients[2, 4], scientific = TRUE)`)

Y- < Mild Leukopenic (P = `r sprintf("%.1e", wbc_Y_Mild$coefficients[2, 4], scientific = TRUE)`)


### Hemoglobin (g/dL)

```{r}
ggplot(clinical_low, aes(x = clust, y = HB)) +
  geom_boxplot() +
  theme_bw() +
    labs(
    title = "Hemoglobin",
    y = "HB (g/dL)",
    x = "Cluster") +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
hb_df <-  clinical_low %>%
  mutate(cl = ifelse(clust == "TET2 bi", "TET2 bi", "Other"))
hb_TET2Bi <- summary(lm(HB ~ cl + SEX + AGE, hb_df, family = "poisson"))
```

TET2bi has higher concentration of hemoglobin (P = `r sprintf("%.1e", hb_TET2Bi$coefficients[2, 4], scientific = TRUE)`) than the other clusters.


## Kariotipos

```{r}
clinical_low$complex <- factor(clinical_low$complex, levels = c("non-complex", "complex"))
kar_events2 <- c("del7", "plus8", "delY", "del20q", "complex")
names(kar_events2) <- kar_events2
kar_tabs <- lapply(kar_events2, function(x) table(clinical_low[[x]], clinical_low$clust))
kar_tabs_lowblasts <- sapply(kar_tabs, function(x) prop.table(x, margin = 2)[2, ])*100

pheatmap(mat = kar_tabs_lowblasts, display_numbers = TRUE)
```

Kariotype aberrations are mostly confined to 8+, del7 and del20q clusters. 

## Mutations

```{r}
sel_muts2 <- c("TET2", "TET2other", "TET2bi", "ASXL1", "DNMT3A", "U2AF1", "RUNX1", "SRSF2", "ZRSR2")
names(sel_muts2) <- sel_muts2
mut_tabs <- lapply(sel_muts2, function(x) table(clinical_low[[x]], clinical_low$clust))

mut_tabs_lowblasts <- sapply(mut_tabs, function(x) prop.table(x, margin = 2)[2, ])*100
pheatmap(mat = mut_tabs_lowblasts, display_numbers = TRUE)

select(clinical_low, sel_muts) %>% colMeans() %>%
  sort(decreasing = TRUE)
```
### TET2 bi

```{r}
clinical_low %>%
  group_by(clust) %>%
  summarize(freq = mean(TET2bi)) %>%
  ggplot(aes(x = clust, y = freq*100)) +
  geom_bar(stat = "identity", color = "#999999") +
  theme_bw() +
  labs(
    title = "TET2 bi-allelic",
    y = "Individuals with mutation (%)",
    x = "Cluster") +
  theme(plot.title = element_text(hjust = 0.5)) 

```
### TET2 mono

```{r}
clinical_low %>%
  group_by(clust) %>%
  summarize(freq = mean(TET2other)) %>%
  ggplot(aes(x = clust, y = freq*100)) +
  geom_bar(stat = "identity", color = "#999999") +
  theme_bw() +
  labs(
    title = "TET2 mono-allelic",
    y = "Individuals with mutation (%)",
    x = "Cluster") +
  theme(plot.title = element_text(hjust = 0.5)) 

```

### ASXL1

```{r}
clinical_low %>%
  group_by(clust) %>%
  summarize(freq = mean(ASXL1)) %>%
  ggplot(aes(x = clust, y = freq*100)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  labs(
    title = "ASXL1 mutations",
    y = "Individuals with mutation (%)",
    x = "Cluster") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylim(c(0, 40))

```

```{r}
asxl1_Ydel20q_other_df <-  clinical_low %>% 
  subset(clust %in% c("del20q", "High Leukopenic", "Mild Leukopenic", "TET2 mono", "Y-")) %>%
  mutate(cl = ifelse(clust %in% c("Y-", "del20q"), "alt", "Other"))
asxl1_Ydel20q_other <- fisher.test(table(asxl1_Ydel20q_other_df$cl,asxl1_Ydel20q_other_df$ASXL1))

asxl1_TET2bi_other_df <-  clinical_low %>% 
  subset(clust %in% c("High Leukopenic", "Mild Leukopenic", "TET2 mono", "TET2 bi", "8+", "del7")) %>%
  mutate(cl = ifelse(clust %in% c("TET2 bi", "8+", "del7"), "High", "Other"))
asxl1_TET2bi_other <- fisher.test(table(asxl1_TET2bi_other_df$cl,asxl1_TET2bi_other_df$ASXL1))
```

ASXL1 frequency depends on subcluster

Y-, del20q < High and Mild Leukopenic, TET2 mono (P = `r sprintf("%.1e", asxl1_Ydel20q_other$p.value, scientific = TRUE)`)

High and Mild Leukopenic, TET2 mono < TET2 bi, 8+, del7 (P = `r sprintf("%.1e", asxl1_TET2bi_other$p.value, scientific = TRUE)`)


### SRSF2

```{r}
clinical_low %>%
  group_by(clust) %>%
  summarize(freq = mean(SRSF2)) %>%
  ggplot(aes(x = clust, y = freq*100)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  labs(
    title = "SRSF2 mutations",
    y = "Individuals with mutation (%)",
    x = "Cluster") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylim(c(0, 40))

```

```{r}
srsf2_groups_df <-  clinical_low %>% 
  mutate(cl = ifelse(clust %in% c("Chromosomal", "TET2 bi", "TET2 mono", "8+", "del7"), "High", "Low"))
srsf2_groups <- fisher.test(table(srsf2_groups_df$cl,srsf2_groups_df$SRSF2))
```

Clusters TET2 mono, TET2 bi, 8+ and del7 have a higher frequency of SRSF2 mutations than Y-, High and Low Leukopenic (P = `r sprintf("%.1e", srsf2_groups$p.value, scientific = TRUE)`)


### U2AF1

```{r}
clinical_low %>%
  group_by(clust) %>%
  summarize(freq = mean(U2AF1)) %>%
  ggplot(aes(x = clust, y = freq*100)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  labs(
    title = "U2AF1 mutations",
    y = "Individuals with mutation (%)",
    x = "Cluster") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylim(c(0, 40))

```

```{r}
u2af1_all_del7del20_df <-  clinical_low %>% 
  mutate(cl = ifelse(clust %in% c("del7", "del20q"), "High", "Other"))
u2af1_all_del7del20 <- fisher.test(table(u2af1_all_del7del20_df$cl,u2af1_all_del7del20_df$U2AF1))
```

del7 and del20q subtypes have a much higher frequency of U2AF1 mutations than the rest (P = `r sprintf("%.1e", u2af1_all_del7del20$p.value, scientific = TRUE)`)

### DNMT3A

```{r}
clinical_low %>%
  group_by(clust) %>%
  summarize(freq = mean(DNMT3A)) %>%
  ggplot(aes(x = clust, y = freq*100)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  labs(
    title = "DNMT3A mutations",
    y = "Individuals with mutation (%)",
    x = "Cluster") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylim(c(0, 40))
```

```{r}
dnmt3a_TET2bi_df <-  clinical_low %>% 
  mutate(cl = ifelse(clust == "TET2 bi", "TET2 bi", "Other"))
dnmt3a_TET2bi <- fisher.test(table(dnmt3a_TET2bi_df$cl,dnmt3a_TET2bi_df$DNMT3A))
```

TET2 bi has much lower frequency of DNMT3A mutations than the other subclusters (P = `r sprintf("%.1e", dnmt3a_TET2bi$p.value, scientific = TRUE)`).

### RUNX1

```{r}
clinical_low %>%
  group_by(clust) %>%
  summarize(freq = mean(RUNX1)) %>%
  ggplot(aes(x = clust, y = freq*100)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  labs(
    title = "RUNX1 mutations",
    y = "Individuals with mutation (%)",
    x = "Cluster") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylim(c(0, 40))

```

```{r}
runx1_YML_HLC_df <-  clinical_low %>% 
  subset(clust %in% c("del20q", "High Leukopenic", "Mild Leukopenic", "Y-")) %>%
  mutate(cl = ifelse(clust %in% c("Y-", "Mild Leukopenic"), "Y-", "Other"))
runx1_YML_HLC <- fisher.test(table(runx1_YML_HLC_df$cl,runx1_YML_HLC_df$RUNX1))

runx1_HLC_TET2_df <-  clinical_low %>% 
  subset(clust %in% c("del20q", "High Leukopenic", "8+", "del7", "TET2 mono", "TET2 bi")) %>%
  mutate(cl = ifelse(clust %in% c("del20q", "High Leukopenic"), "Low", "Other"))
runx1_HLC_TET2 <- fisher.test(table(runx1_HLC_TET2_df$cl,runx1_HLC_TET2_df$RUNX1))

runx1_ML_TET2_df <-  clinical_low %>% 
  subset(clust %in% c("Y-", "Mild Leukopenic", "8+", "del7", "TET2 mono", "TET2 bi")) %>%
  mutate(cl = ifelse(clust %in% c("Y-", "Mild Leukopenic"), "Low", "Other"))
runx1_ML_TET2 <- fisher.test(table(runx1_ML_TET2_df$cl,runx1_ML_TET2_df$RUNX1))


```

RUNX1 frequency depends on subcluster

Y-, Mild Leukopenic < High Leukopenic, del20q (P = `r sprintf("%.1e", runx1_YML_HLC$p.value, scientific = TRUE)`)

High Leukopenic, del20q < TET2 bi, TET2 mono, 8+, del7 (P = `r sprintf("%.1e", runx1_HLC_TET2$p.value, scientific = TRUE)`)

Y-, Mild Leukopenic < TET2 bi, TET2 mono, 8+, del7 (P = `r sprintf("%.1e", runx1_ML_TET2$p.value, scientific = TRUE)`)


## IPSSM 

```{r}
clinical_IPSSM <- mutate(clinical_low, 
                         IPSSM_col = ifelse(IPSSM %in% c("High", "Very-High"), "High/Very-High", 
                                            ifelse(grepl("Moderate", IPSSM), "Moderate",
                                                     ifelse(IPSSM == "Low", "Low", "Very-Low"))),
                         IPSSM_col = factor(IPSSM_col, levels = c("Very-Low", "Low", "Moderate", "High/Very-High")))
IPSSM_cats <- levels(clinical_IPSSM$IPSSM_col)
names(IPSSM_cats) <- IPSSM_cats
```



```{r}
clinical_low %>%
  filter(!is.na(IPSSM)) %>%
  group_by(clust, IPSSM) %>%
  summarize(N = n()) %>%
  group_by(clust) %>%
  mutate(Freq = N/sum(N)) %>%
  ggplot(aes(x = clust, y = Freq*100, fill = fct_rev(IPSSM))) +
  geom_bar(stat = "identity") +
  theme_bw() +
  scale_fill_manual(values = c("#d73027", "#f46d43", "#fdae61", "#fee08b", "#66bd63", "#2ca25f")) +
  labs(
    title = "IPSSM classification",
    y = "Proportion of individuals (%)",
    x = "Cluster",
    fill = "IPSSM Category") +
  theme(plot.title = element_text(hjust = 0.5))
```
```{r}
clinical_IPSSM %>%
  filter(!is.na(IPSSM_col)) %>%
  group_by(clust, IPSSM_col) %>%
  summarize(N = n()) %>%
  group_by(clust) %>%
  mutate(Freq = N/sum(N)) %>%
  ggplot(aes(x = clust, y = Freq*100, fill = fct_rev(IPSSM_col))) +
  geom_bar(stat = "identity") +
  theme_bw() +
  scale_fill_manual(values = c("#d73027", "#fee08b", "#66bd63", "#2ca25f")) +
  labs(
    title = "IPSSM collapsed classification",
    y = "Proportion of individuals (%)",
    x = "Cluster",
    fill = "IPSSM Category") +
  theme(plot.title = element_text(hjust = 0.5))
```


```{r}
ipssm_MLY_TET2HL_df <-  clinical_IPSSM %>% 
  subset(clust %in% c("Mild Leukopenic", "High Leukopenic", "TET2 bi", "TET2 mono", "Y-", "del20q")) %>%
  mutate(cl = ifelse(clust %in% c("Mild Leukopenic", "Y-"), "Low", "Other"))
ipssm_MLY_TET2HL <- chisq.test(table(ipssm_MLY_TET2HL_df$cl,ipssm_MLY_TET2HL_df$IPSSM_col))

ipssm_TET2HL_Chr_df <-  clinical_IPSSM %>% 
  subset(clust %in% c("8+", "del7", "High Leukopenic", "TET2 bi", "TET2 mono")) %>%
  mutate(cl = ifelse(clust %in% c("8+", "del7"), "Chromosomal", "Other"))
ipssm_TET2HL_Chr <- chisq.test(table(ipssm_TET2HL_Chr_df$cl,ipssm_TET2HL_Chr_df$IPSSM_col))
```

The sub-clusters have different distribution of IPSSM scores:

Y-, Mild Leukopenic < High Leukopenic, TET2 bi, TET2 mono, del20q (P = `r sprintf("%.1e", ipssm_MLY_TET2HL$p.value, scientific = TRUE)`).


High Leukopenic, TET2 bi, TET2 mono, del20q < 8+, del7 (P = `r sprintf("%.1e", ipssm_TET2HL_Chr$p.value, scientific = TRUE)`).

## Survival


```{r}
group_cols <- c("#000000", "#999999", "#E69F00",  
  "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00"   
)
surv_low <- survfit(formula = Surv(OS_YEARS,OS_STATUS) ~ clust, clinical_low) %>%
    ggsurvplot(data = clinical_low, surv.median.line = "hv",
               palette = group_cols) +
  xlab("Time (Years)") +
  ggtitle("Overall Survival") 

surv_low$plot  +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
os_Ch_TET2bHL_df <- filter(clinical_low, 
                      clust %in% c("8+", "del7", "del20q", "TET2 bi", "High Leukopenic")) %>%
  mutate(cl = ifelse(clust %in% c("8+", "del7"), "Chromosomal", "Other"))
os_Ch_TET2bHL <- summary(coxph(Surv(OS_YEARS,OS_STATUS) ~ cl + AGE, os_Ch_TET2bHL_df))

os_TET2bHL_TET2mML_df <- filter(clinical_low, 
                      clust %in% c("TET2 mono", "Mild Leukopenic", "TET2 bi", "High Leukopenic", "Y-", "del20q")) %>%
  mutate(cl = ifelse(clust %in% c("TET2 mono", "Mild Leukopenic", "Y-"), "Low", "Other"))
os_TET2bHL_TET2mML <- summary(coxph(Surv(OS_YEARS,OS_STATUS) ~ cl + AGE, os_TET2bHL_TET2mML_df))
```

The subclusters have different survivals:

8+, del7 < TET2 bi, High Leukopenic, del20q (P = `r sprintf("%.1e", os_Ch_TET2bHL$coefficients[1, 5], scientific = TRUE)`)

TET2 bi, High Leukopenic, del20q < TET2 mono, Mild Leukopenic, Y- (P = `r sprintf("%.1e", os_TET2bHL_TET2mML$coefficients[1, 5], scientific = TRUE)`)

### Survival vs IPSSM

```{r}
clusts <- levels(clinical_low$clust)
survs <- lapply(IPSSM_cats, function(cat){
  df <- filter(clinical_IPSSM, IPSSM_col == cat)
  n_group <- df %>% 
    group_by(clust) %>%
    summarize(n = n())
  sel_clusts <- as.character(filter(n_group, n >= 10)$clust)
  df <- filter(df, clust %in% sel_clusts)
  p <- survfit(formula = Surv(OS_YEARS,OS_STATUS) ~ clust, df) %>%
    ggsurvplot(data = df, surv.median.line = "hv",
               palette = group_cols[which(clusts %in% sel_clusts)]) +
    ggtitle(cat)
  p$plot
})
survs
```

```{r}
os_vl_TET2_df <- clinical_IPSSM %>%
  filter(IPSSM_col == "Very-Low" & clust %in% c("Y-", "Mild Leukopenic", "High Leukopenic", "TET2 bi", "TET2 mono")) %>%
  mutate(cl = ifelse(clust %in% c("TET2 bi", "TET2 mono"), "A", "B"))
os_vl_TET2 <- summary(coxph(Surv(OS_YEARS,OS_STATUS) ~ cl + AGE, os_vl_TET2_df))
```

In patients with very-low IPSSM, TET2 bi and TET2 mono have lower survival, though it is not significant (P = `r sprintf("%.1e", os_vl_TET2$coefficients[1, 5], scientific = TRUE)`).

```{r}
os_l_TET2_df <- clinical_IPSSM %>%
  filter(IPSSM_col == "Low" & clust %in% c("del20q", "Mild Leukopenic", "High Leukopenic", "TET2 bi", "TET2 mono")) %>%
  mutate(cl = ifelse(clust %in% c("TET2 mono"), "A", "B"))
os_l_TET2 <- summary(coxph(Surv(OS_YEARS,OS_STATUS) ~ cl + AGE, os_l_TET2_df))

os_l_Y_df <- clinical_IPSSM %>%
  filter(IPSSM_col == "Low" & clust %in% c("del20q", "Mild Leukopenic", "High Leukopenic", "TET2 bi", "Y-")) %>%
  mutate(cl = ifelse(clust %in% c("Y-"), "A", "B"))
os_l_Y <- summary(coxph(Surv(OS_YEARS,OS_STATUS) ~ cl + AGE, os_l_Y_df))


os_l_TET2_Y_df <- clinical_IPSSM %>%
  filter(IPSSM_col == "Low" & clust %in% c("TET2 mono", "Y-")) %>%
  mutate(cl = ifelse(clust %in% c("Y-"), "A", "B"))
os_l_TET2_Y <- summary(coxph(Surv(OS_YEARS,OS_STATUS) ~ cl + AGE, os_l_TET2_Y_df))

```


In patients with IPSSM low, TET2 mono patients had the best survival, while Y- the worst. Although the difference in survival between TET2 mono and Y- was statistically significant (P = `r sprintf("%.1e", os_l_TET2_Y$coefficients[1, 5], scientific = TRUE)`), the differenes between TET2 mono and the rest (P = `r sprintf("%.1e", os_l_TET2$coefficients[1, 5], scientific = TRUE)`) or Y- and the rest (P = `r sprintf("%.1e", os_l_Y$coefficients[1, 5], scientific = TRUE)`) were not. 


```{r}
os_m_TET2b8_df <- clinical_IPSSM %>%
  filter(IPSSM_col == "Moderate" & clust %in% c("8+", "Mild Leukopenic", "High Leukopenic", "TET2 bi", "TET2 mono")) %>%
  mutate(cl = ifelse(clust %in% c("TET2 bi", "8+"), "A", "B"))
os_m_TET2b8 <- summary(coxph(Surv(OS_YEARS,OS_STATUS) ~ cl + AGE, os_m_TET2b8_df))
```

In patients with moderate IPSSM, TET2 bi and 8+ have lower survival (P = `r sprintf("%.1e", os_m_TET2b8$coefficients[1, 5], scientific = TRUE)`).


```{r}
os_h_TET2bHL7_df <- clinical_IPSSM %>%
  filter(IPSSM_col == "High/Very-High" & clust %in% c("8+", "Mild Leukopenic", "High Leukopenic", "TET2 bi", "TET2 mono", "del7")) %>%
  mutate(cl = ifelse(clust %in% c("TET2 bi", "High Leukopenic", "del7"), "A", "B"))
os_h_TET2bHL7 <- summary(coxph(Surv(OS_YEARS,OS_STATUS) ~ cl + AGE, os_h_TET2bHL7_df))
```

In patients with high or very-high IPSSM, TET2 bi, high leukopenic and 7- have lower survival (P = `r sprintf("%.1e", os_h_TET2bHL7$coefficients[1, 5], scientific = TRUE)`).

### Survival vs IPSSM per cluster


```{r}
main_clusts <- c("High Leukopenic", "Mild Leukopenic", "TET2 mono", "TET2 bi")
names(main_clusts) <- main_clusts
ipssm_cols <- c("#E0F3DB", "#A8DDB5", "#43A2CA","#0868AC")

survs2 <- lapply(main_clusts, function(cl){
  df <- filter(clinical_IPSSM, clust == cl)
  p <- survfit(formula = Surv(OS_YEARS,OS_STATUS) ~ IPSSM_col, df) %>%
    ggsurvplot(data = df, surv.median.line = "hv",
               palette = ipssm_cols) +
    ggtitle(cl)
  p$plot
})
survs2
```

```{r}
clinical_all_IPSSM <- mutate(clinical, 
                         IPSSM_col = ifelse(IPSSM %in% c("High", "Very-High"), "High/Very-High", 
                                            ifelse(grepl("Moderate", IPSSM), "Moderate",
                                                     ifelse(IPSSM == "Low", "Low", "Very-Low"))),
                         IPSSM_col = factor(IPSSM_col, levels = c("Very-Low", "Low", "Moderate", "High/Very-High")))
```

```{r}
computeIPSSM_HRs <- function(df){
  
  inds <- seq_len(length(IPSSM_cats) - 1)
  HRs <- lapply(inds, function(i){
    
    cats <- IPSSM_cats[c(i, i+1)]
    
    tmp <- subset(df, IPSSM_col %in% cats) %>%
      mutate(IPSSM_col = droplevels(IPSSM_col))
    
    mod <- summary(coxph(Surv(OS_YEARS,OS_STATUS) ~ IPSSM_col + AGE, tmp))
    out <- mod$conf.int[1, -2]
    out
  })
  res <- Reduce(rbind, HRs) %>%
    as_tibble() %>%
    mutate(Comparison = c("Very-Low vs Low", "Low vs Moderate", "Moderate vs High/Very-High"))
  colnames(res) <- c("HR", "L_95", "H_95", "Comparison")
  res
}

clust_hrs <- lapply(main_clusts, function(cl){
  
  tab <- computeIPSSM_HRs(filter(clinical_IPSSM, clust == cl)) %>%
    mutate(Cluster = cl)
  tab
}) %>%
  Reduce(f = rbind)

all_hrs <- rbind(clust_hrs, 
        computeIPSSM_HRs(clinical_IPSSM) %>%
          mutate(Cluster = "Low Blasts"),
        computeIPSSM_HRs(clinical_all_IPSSM) %>%
          mutate(Cluster = "IPSS-Mol cohort")) %>%
  mutate(Comparison = factor(Comparison, levels = c("Very-Low vs Low", "Low vs Moderate", "Moderate vs High/Very-High")),
         Cluster = factor(Cluster, levels = c(main_clusts, "Low Blasts", "IPSS-Mol cohort")))

```

```{r}
ggplot(all_hrs, aes(x = Cluster, y = HR, fill = Cluster)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  geom_errorbar(aes(ymin = L_95, ymax = H_95), width = 0.2) +
  theme_bw() +
  facet_grid(~ Comparison) +
  scale_y_log10() +
  scale_fill_manual(values = c(group_cols[1:4], "white", "purple")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))


```
# Mutations

## Survival vs mutations

```{r}
muts_df <- clinical_low %>%
  mutate(ASXL1 = ifelse(ASXL1 == 1, "mutated", "WT"),
         SRSF2 = ifelse(SRSF2 == 1, "mutated", "WT"),
         cl = ifelse(clust == "TET2 bi", "1.TET2 bi", "2.Other")) 

```



```{r}
survfit(formula = Surv(OS_YEARS,OS_STATUS) ~ ASXL1, muts_df) %>%
    ggsurvplot_facet(data = muts_df, surv.median.line = "hv", 
              facet.by = "cl")
```
```{r}
summary(coxph(Surv(OS_YEARS,OS_STATUS) ~ cl*ASXL1 + AGE, muts_df))
```


```{r}
survfit(formula = Surv(OS_YEARS,OS_STATUS) ~ SRSF2, muts_df) %>%
    ggsurvplot_facet(data = muts_df, surv.median.line = "hv", 
              facet.by = "cl")
```


```{r}
summary(coxph(Surv(OS_YEARS,OS_STATUS) ~ cl*SRSF2  + AGE, muts_df))
```

```{r}
summary(coxph(Surv(OS_YEARS,OS_STATUS) ~ cl*ASXL1 + AGE, muts_df, subset = clust %in% main_clusts))
summary(coxph(Surv(OS_YEARS,OS_STATUS) ~ cl*ASXL1 + AGE, muts_df))
```

```{r}
summary(coxph(Surv(OS_YEARS,OS_STATUS) ~ cl*RUNX1 + AGE, muts_df))
```

```{r}
summary(coxph(Surv(OS_YEARS,OS_STATUS) ~ cl*EZH2 + AGE, muts_df, subset = clust %in% main_clusts))
summary(coxph(Surv(OS_YEARS,OS_STATUS) ~ cl*EZH2 + AGE, muts_df))
```


```{r}
TET2bi <- subset(clinical_low, clust == "TET2 bi")
summary(coxph(Surv(OS_YEARS,OS_STATUS) ~ AGE  + BM_BLAST + logPLT + HB + ASXL1 + RUNX1 + SRSF2, TET2bi))

TET2bi_no <- subset(clinical_low, clust != "TET2 bi")
summary(coxph(Surv(OS_YEARS,OS_STATUS) ~ AGE  + BM_BLAST + logPLT + HB + ASXL1 + RUNX1 + SRSF2, TET2bi_no))


summary(coxph(Surv(OS_YEARS,OS_STATUS) ~ AGE  + BM_BLAST + log(PLT) + HB + ASXL1 + RUNX1 + SRSF2, clinical))
```



## Co-mutations

### ASXL1 vs RUNX1

```{r}
table(clinical_low$ASXL1, clinical_low$RUNX1)
fisher.test(table(clinical_low$ASXL1, clinical_low$RUNX1))
lapply(main_clusts, function(x){
  tmp <- subset(clinical_low, clust == x)
  list(tab =  table(tmp$ASXL1, tmp$RUNX1),
       test = fisher.test(table(tmp$ASXL1, tmp$RUNX1)))
})
```

### SRSF2 vs RUNX1

```{r}
table(clinical_low$SRSF2, clinical_low$RUNX1)
fisher.test(table(clinical_low$SRSF2, clinical_low$RUNX1))
lapply(main_clusts, function(x){
  tmp <- subset(clinical_low, clust == x)
  list(tab =  table(tmp$SRSF2, tmp$RUNX1),
       test = fisher.test(table(tmp$SRSF2, tmp$RUNX1)))
})
```
### ASXL1 vs DNMT3A

```{r}
table(clinical_low$ASXL1, clinical_low$DNMT3A)
fisher.test(table(clinical_low$ASXL1, clinical_low$DNMT3A))
lapply(main_clusts, function(x){
  tmp <- subset(clinical_low, clust == x)
  list(tab =  table(tmp$ASXL1, tmp$DNMT3A),
       test = fisher.test(table(tmp$ASXL1, tmp$DNMT3A)))
})
```

### ASXL1 vs SRSF2


```{r}
table(clinical_low$ASXL1, clinical_low$SRSF2)
fisher.test(table(clinical_low$ASXL1, clinical_low$SRSF2))
lapply(main_clusts, function(x){
  tmp <- subset(clinical_low, clust == x)
  list(tab =  table(tmp$ASXL1, tmp$SRSF2),
       test = fisher.test(table(tmp$ASXL1, tmp$SRSF2)))
})
```

