---
title: "Mutation Data Exploration"
author: "Carlos Ruiz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

In this document, I will explore the clinical variables present in the IPPSmol dataset.

# Data loading

```{r}
library(tidyverse)
library(GGally)
library(survival)
library(survminer)
library(cowplot)
```

```{r}
mutation <- read_tsv("../data/IPSSMol/df_mut.tsv")
```

```{r}
colnames(mutation)
```

The dataset contains `r nrow(mutation)` individuals and `r ncol(mutation)` variables. All the variables are binary.

# Data exploration

```{r}
mut_sing <- mutation[, 1:127] ## Remove columns having subtypes of TP53 and TET2
```


## Mutations per individual

```{r}
data.frame(n = rowSums(mut_sing[, -1])) %>%
  ggplot(aes(x = n)) +
  geom_histogram(binwidth = 1) +
  xlab("Mutations per individual") +
  theme_bw()

data.frame(n = rowSums(mut_sing[, -1])) %>%
    mutate(intervals = cut(n, 
          breaks = c(-1, 0, 1, 2, 4, 20), 
           labels = c("0", "1", "2", "3-4", "4+"))) %>%
    count(intervals) %>%
    mutate(perc = round(n/sum(n)*100, 1))

```

Most patients (~90%) have mutations, with half of the patients having 3 or more mutations. 

## Mutations frequency

```{r}
data.frame(n = colMeans(mut_sing[, -1])) %>%
  ggplot(aes(x = n*100)) +
  geom_histogram(binwidth = 1) +
  xlab("Mutation frequency") +
  theme_bw()

data.frame(n = colMeans(mut_sing[, -1])) %>%
    mutate(intervals = cut(n*100, 
          breaks = c(-1, 1, 5, 10, 20, 50), 
           labels = c("<1%", "1-5%", "5-10%", "10-20%", "20-50%"))) %>%
    count(intervals) %>%
    mutate(perc = round(n/sum(n)*100, 1))

mut_sing %>% 
  pivot_longer(2:127, names_to = c("Gene")) %>%
  group_by(Gene) %>%
  summarise(n = sum(value),
            prop = round(mean(value)*100, 1)) %>%
  arrange(desc(n))
```

Most mutations were very rare, being present in less than 1% of the patients. Other mutations are much frequent, such as TET2, ASXL1 or SF3B1. 



